Seg_sites_rarefaction

### 2023-12-20 manuscript/ms.Rmd
Introduction draft added

# Introduction
<!-- Outline: -->

@@

# Methods
Outline:

* Background on methods for private alleles, richness
* Solution for segregating sites
  - Derivation
  - Can account for both different sample sizes and missingness across loci and populations.
* Simulations
  - Sampling approach, missingness, etc

<!-- https://link.springer.com/article/10.1023/B:COGE.0000041021.91777.1a#citeas -->

@@

However, HWE is often not a desirable assumption to make. Even if filtering is employed to remove loci which do not conform to HWE, the degree of conformity, and thus the degree of statistical bias in estimating $P({S_{j}})$, typically varies somewhat between populations.

To remedy this, I propose the following estimator of $P({S_{j}})$:
$$P({S_{j}}) = 1 - \sum_{k = 1}^{h}\frac{\binom{n_{kj}}{\gamma}}{\binom{n_{j}}{\gamma}}$$
where the $P({S_{j}})$ is given by the probability of exclusively drawing any $k$ of $h$ possible homozygote genotypes in population $j$ given $\gamma$ independent sampled *genotypes* (not *gene copies*) from the pool of observed genotypes. Here, $n_{kj}$ is the number of observed $k$ genotypes in population $j$ where the total number of observed genotypes of all types is $n_{j}$. Note that $n_{j}$ and $\gamma$ will be half the value of their equivalents $N_{j}$ and $g$ for diploid species, one third for triploids, and so on.

@@

Confidence and prediction intervals can then be derived using standard approaches for the sum of random, independent Bernoulli trials. When $Q$ is large, for example, the distribution of $E(N_{S})$ should be approach normal and confidence and prediction intervals can be derived using standard normal approximation. In contrast, confidence intervals for individual $P({S_{jq}})$ values can be achieved via standard approaches for single Bernoulli samples of size $\gamma$.

## Emperical Validation
To validate equations 4 and 5, I simulated genotypic data for two populations with sizes 100 and 1000, each with 100 bi-allelic loci with minor allele frequencies spaced equally between 0.01 and 0.1. I added missing data to each population assigning each locus a missing data rate $R_{mq}$ from a uniform distribution such that $R_{mq} \sim U(0,.3)$, ensuring that overall allele frequencies in each population were maintained. I then used the methods described above to estimate $P({S_{jq}})$ and $E(N_{S})$ and their variances given $\gamma = 30$. I used the "asymptotic" method from the R package `binom` [@rajBinomBinomialConfidence2022] to calculate 95% confidence intervals for each $P({S_{jq}})$ and normal approximation to calculate 95% confidence and prediction intervals for $E(N_{S})$.

For comparison, I also conducted 10,000 random draws of size $\gamma$ from each loci in each population, then calculated $P({S_{jq}})$ and its variance for each locus empirically and $E(N_{S})$ by summing across all 100 loci for each set of draws. I likewise calculated the variance of $E(N_{S})$ directly across all sets of random draws.

# Results
The methods described here for calculating $P({S_{j}})$ and $E(N_{S})$ performed well. 95% confidence intervals for simulated loci overlapped the simulated mean in

>>>>

# Introduction

Estimating the degree of genetic diversity has long been a critically important and widespread practice in conservation genetics [@dewoodyLongstandingSignificanceGenetic2021]. Genetic diversity, or the diversity alleles, genotypes, and segregating loci within a population [@frankhamIntroductionConservationGenetics2002], is the both the basic foundation upon which selection acts and thus in part controls population adaptability [@Kardos2021], and is strongly correlated with fitness both at the individual and population level, even when no obvious functional ties are known [reviewed in @dewoodyLongstandingSignificanceGenetic2021].

Estimators of population genetic diversity are themselves extremely diverse, but can essentially be split into two categories depending on if they are allele frequency dependent or independent. Allele frequency dependent statistics, such as expected heterozygosity ($H_{e}$), the F-statistics ($F_{IS}$, $F_{ST}$, and $F_{IT}$)[@Weir1984], nucleotide diversity ($\pi$), and Tajima's $\theta$ [@Tajima1989] all rely principally on the frequency of the alleles across any surveyed loci; whereas allele frequency independent statistics, such as Watterson's $\theta$ [@wattersonNumberSegregatingSites1975], counts of alleles per loci, the number of observed segregating sites, and (to some degree) private allele counts do not. This is not a trivial difference:  evolutionary forces do not act equally on the number of loci/alleles in a population and the frequency of those alleles. Tajima's D, for example, is a powerful and broadly used statistic fundamentally based on the difference between Tajima's $\theta$ and Watterson's $\theta$ can be used to both detect selection and population demographic changes [@Tajima1989] due to the way that those forces act on the balance of allele frequencies within populations.

The average allele count per locus (which is usually corrected to allelic richness [@kalinowskiCountingAllelesRarefaction2004]) has historically been one of the most prevalent allele frequency independent measures of genetic diversity. While this is a particularly useful measure in microsatellite and other datasets utilizing heavily poly-allelic markers, it is less so in single-nucleotide polymorphism (SNP) datasets. SNP datasets are usually primarily biallelic, and thus the count of the number of alleles per locus in such datasets is less variable across populations and therefore less interesting. Counts of the number of segregating sites per population are a useful alternative in SNP datasets given the large number of possible segregating sites that these datasets often query [@?CITE].

However, raw counts of the number of segregating sites per population are problematic whenever sample sizes are not equal across the populations under comparison at all loci. This can occur either directly due to unequal numbers of individuals or  indirectly due to unqual proportions of missing data across populations. This problem also effects estimates of allele counts per locus and private allele numbers, but corrections for both which use rarefaction to estimate those parameters under a common sample size are well developed [@kalinowskiCountingAllelesRarefaction2004].

Here we present rarefaction-corrected estimators for the probability that any given loci would be observed to be segregating under a reduced sample size and the expected total number of segregating sites across all loci within a population. We show that these estimators are highly accurate via comparison to simulated re-sampling using simulated data. These estimators are currently implemented and available in the `snpR` R package via the function `calc_seg_sites` [@hemstromSnpRUserFriendly2023], where they can be automatically run for all populations either with automatic, per-locus optimized control of target sample size or globally via manual control.

<!-- Outline: -->

@@

# Methods
<!-- https://link.springer.com/article/10.1023/B:COGE.0000041021.91777.1a#citeas -->

@@

However, HWE is often not a desirable assumption to make. Even if filtering is employed to remove loci which do not conform to HWE, the degree of conformity, and thus the degree of statistical bias in estimating $P({S_{j}})$, typically varies somewhat between populations. For example, in a sample of 100 genotypes with a minor allele frequency of 0.05, only five minor alleles are expected and two out of three possible combinations of minor homozygotes and heterozygotes that produce that frequency will not deviate from HWE at $\alpha = 0.05$ according to an exact test [@Wigginton2005]. However, re-sampling these to, say, ten genoytpes should will produce quite different $P(S_{j})$ (roughly 0.7, 0.6, and 0.5, for purely heterozygotes, one homozygote and three heterozygotes, and two homozygotes and one heterozygote, respectively)  as we will see below.

To remedy these problems, I propose the following estimator of $P({S_{j}})$:
$$P({S_{j}}) = 1 - \sum_{k = 1}^{h}\frac{\binom{n_{kj}}{\gamma}}{\binom{n_{j}}{\gamma}}$$
where the $P({S_{j}})$ is given by the probability of exclusively drawing any $k$ of $h$ possible homozygote genotypes in population $j$ given $\gamma$ independent sampled *genotypes* (not *gene copies*) from the pool of observed genotypes. Here, $n_{kj}$ is the number of observed homozygote genotypes of type $k$ in population $j$ and $n_{j}$ is the total number of observed genotypes of all types. Note that $n_{j}$ and $\gamma$ will be half the value of their equivalents $N_{j}$ and $g$ for diploid species, one third for triploids, and so on.

@@

Confidence and prediction intervals can then be derived using standard approaches for the sum of random, independent Bernoulli trials. When $Q$ is large, for example, the distribution of $E(N_{S})$ should be approach normal and confidence and prediction intervals can be derived using standard normal approximation using the equations $$CI_{N_{S}} = E(N_{S}) \pm Z\sqrt{\frac{\sigma_{E(n_{S})}^{2}}{Q}}$$ $$PI_{N_{S}} = E(N_{S}) \pm Z\sqrt{\sigma_{E(n_{S})}^{2}(1+(1/Q))}$$
where $Z$ is given by the normal quantile function $Z = Q_{X}(1-\alpha)$ with $\mu = 0$ and $\sigma = 1$ for a desired confidence level $\alpha$.

## Emperical Validation
To validate equations 4 and 5, I simulated genotypic data for two populations with sizes 100 and 1000, each with 100 bi-allelic loci with minor allele frequencies spaced equally between 0.01 and 0.1. I added missing data to each population assigning each locus a missing data rate $R_{mq}$ from a uniform distribution such that $R_{mq} \sim U(0,.3)$, ensuring that overall allele frequencies in each population were maintained. I then used the methods described above to estimate $P({S_{jq}})$ and $E(N_{S})$ and their variances given $\gamma = 30$. For comparison, I also conducted 10,000 random draws of size $\gamma$ from each loci in each population, then calculated $P({S_{jq}})$ and its variance for each locus empirically and $E(N_{S})$ by summing across all 100 loci for each set of draws. I likewise calculated the variance of $N_{S}$ directly across all sets of random draws.

To compare my calculated estimates to the empirical simulations, I used the implementation of the Agresti-Coull [@agrestiApproximateBetterExact1998] method from the R package `binom` [@dorai-rajBinomBinomialConfidence2022] to calculate 95% confidence intervals for each empirical $P({S_{jq}})$. I also used normal approximation to calculate 95% confidence and prediction intervals for $E(N_{S})$ and $N_{S}$ using $\sigma_{E(n_{S})}^{2}$ and the variance directly observed from the simulations, respectively.

# Results
The methods described here for calculating $P({S_{j}})$ and $E(N_{S})$ performed well. Individual $P({S_{jq}})$ values for each locus were within the confidence intervals derived from the simulated values for 97 and 96% of loci from the $n = 100$ and $n = 1000$ populations, respectively (Figure 1). Note that variation along the generally correlated minor allele frequency/$P({S_{jq}})$ axes is due to variations in genotype frequencies in the simulated data for a given minor allele frequency. $P({S_{jq}})$ values estimated using equation 4 account for this adequately.

$E(N_{S})$ values estimated with equation 5 were similarly accurate and were within the 95% confidence intervals produced using the simulated $N_{S}$ for both population sizes (Figure 2). Likewise, the 95% prediction intervals calculated using $\sigma_{E(n_{S})}^{2}$ via equation 7 contained 96.1 and 95.3% of the simulated $N_{S}$ values for $n = 100$ and $n = 1000$, respectively (Figure 2).

# Discussion

# References

### 2023-12-20 scripts/seg_sites_rarefaction.Rmd
Introduction draft added

```
# Define parameters
```{r}
# Generate some example data:
## define parameters
seed <- 231232
allele_frequencies <- seq(.001, .1, by = .001) # allele frequencies
n1 <- 2000 # p1 gene copy number
n2 <- 200 # p2 gene copy number
k <- 30 # number to rarefact to
mcmcs <- 10000 # number of mcmc draws
max_missing <- 30 # maximum percentage of missing data in a locus, from a uniform dist

@@

# populate
## function to replace minors
populate <- function(g, missing, minors){
  g[missing] <- NA
  g[-missing][minors] <- 1
  return(g)
}
## do for every locus
for(i in 1:nal){ 
  g1[i,] <- populate(g1[i,], ms1[[i]], ma1[[i]])
  g2[i,] <- populate(g2[i,], ms2[[i]], ma2[[i]])
}

@@

# taken from the binom package, asymptotic method
confint_pb <- function(p, n, alpha){
  alpha <- rep(alpha, length = length(p))
  alpha2 <- 0.5 * alpha
  z <- qnorm(1 - alpha2)
  z2 <- z * z
  se <- sqrt(p * (1 - p)/n)
  lcl <- p - z * se
  ucl <- p + z * se

  lcl[lcl < 0] <- 0
  ucl[ucl > 1] <- 1
  return(data.frame(lower = lcl, upper = ucl))
}

@@

vseg1 <- varseg(pseg1)
vseg2 <- varseg(pseg2)
ciseg1 <- confint_pb(pseg1, k, .05)
ciseg2 <- confint_pb(pseg2, k, .05)

@@

mpseg1 <- binom::binom.confint(rowSums(seg1), rep(mcmcs, nrow(seg1)), 
                               method = "exact")[,c("mean", "lower", "upper")]
mpseg2 <- binom::binom.confint(rowSums(seg2), rep(mcmcs, nrow(seg1)),
                               method = "exact")[,c("mean", "lower", "upper")]

@@

# results
```{r}
# individual loci
pseg <- rbind(data.frame(pop = paste0("n = ", n1/2), p = pseg1, snp = 1:nal, af = allele_frequencies),
              data.frame(pop = paste0("n = ", n2/2), p = pseg2, snp = 1:nal, af = allele_frequencies))

@@

ggplot(seg[seg$source == "Mathematical",],
       aes(x = af, y = p)) +
  geom_point(aes(color = in_ci)) +
  geom_errorbar(data = cis, aes(ymax = upper, ymin = lower)) +
  facet_wrap(~pop) +
  theme_bw()

@@

msum_tab <- data.frame(nseg = c(mean(msum1), mean(msum2)),
                       lower = c(quantile(msum1, .05), quantile(msum2, .05)),
                       upper = c(quantile(msum1, .95), quantile(msum2, .95)),
                       pop = c(paste0("n = ", n1/2), paste0("n = ", n2/2)))

sum_tab <- tapply(seg[seg$source == "Mathematical",]$p, seg[seg$source == "Mathematical",]$pop, sum)
sum_tab <- as.data.frame(sum_tab)
sum_tab$pop <- rownames(sum_tab)
colnames(sum_tab)[1] <- "nseg"

all_mcmc_sums <- rbind(cbind.data.frame(nseg = msum1, pop = paste0("n = ", n1/2)),
                   cbind.data.frame(nseg = msum2, pop = paste0("n = ", n2/2)))

ggplot(sum_tab, aes(x = pop, y = nseg)) + 
  geom_point(color = "red") +
  geom_errorbar(data = msum_tab, aes(ymax = upper, ymin = lower, x= pop), inherit.aes = FALSE) +
  geom_jitter(data = all_mcmc_sums, height = 0, alpha = .005) +
  theme_bw() +
  scale_y_continuous(limits = c(0, nal))

tapply(seg[seg$source == "Mathematical",]$in_ci, seg[seg$source == "Mathematical",]$pop, mean, na.rm = TRUE)

@@

# variances, overall
overall_vars <- rbind(cbind.data.frame(estimated = sum(vseg1),
                                       observed = var(msum1),
                                       pop = paste0("n = ", n1/2)),
                      cbind.data.frame(estimated = sum(vseg2),
                                       observed = var(msum2),
                                       pop = paste0("n = ", n2/2)))

# estimated PI -- normal approx works well
pi1 <- qnorm(.975)*sqrt(sum(vseg1))*sqrt(1 + (1/nal))
pi2 <- qnorm(.975)*sqrt(sum(vseg2))*sqrt(1 + (1/nal))
pi <- data.frame(pop = paste0("n = ", n1/2), 
                 lower = sum(pseg1) - pi1,
                 upper = sum(pseg1) + pi1)
pi <- rbind(pi, data.frame(pop = paste0("n = ", n2/2), 
                 lower = sum(pseg2) - pi2,
                 upper = sum(pseg2) + pi2))

all_mcmc_sums <- as.data.table(all_mcmc_sums)
all_mcmc_sums[pop == pi$pop[1], 
              in_pi := nseg <= pi$upper[1] & 
                nseg >= pi$lower[1]]
all_mcmc_sums[pop == pi$pop[2], 
              in_pi := nseg <= pi$upper[2] & 
                nseg >= pi$lower[2]]

ggplot(all_mcmc_sums, aes(x = pop, y = nseg, color = in_pi)) +
  geom_jitter(height = 0, alpha = .5) +
  geom_errorbar(data = pi, aes(ymax = upper, ymin = lower, x = pop), inherit.aes = FALSE)

tapply(all_mcmc_sums$in_pi, all_mcmc_sums$pop, mean)
```

>>>>

```
# Define parameters
```{r}
# Generate some example data:
## define parameters
seed <- 231233
allele_frequencies <- seq(.001, .1, by = .001) # allele frequencies
n1 <- 2000 # p1 gene copy number
n2 <- 200 # p2 gene copy number
k <- 30 # number to rarefact to
mcmcs <- 10000 # number of mcmc draws
max_missing <- 30 # maximum percentage of missing data in a locus, from a uniform dist

@@

# populate
## function to replace minors
populate <- function(g, missing, minors){
  if(length(missing) > 0){
    g[missing] <- NA
    g[-missing][minors] <- 1
  }
  else{
    g[minors] <- 1
  }
  return(g)
}
## do for every locus
for(i in 1:nal){ 
  g1[i,] <- populate(g1[i,], ms1[[i]], ma1[[i]])
  g2[i,] <- populate(g2[i,], ms2[[i]], ma2[[i]])
}

@@

@@

vseg1 <- varseg(pseg1)
vseg2 <- varseg(pseg2)

@@

mpseg1 <- binom::binom.confint(rowSums(seg1), rep(mcmcs, nrow(seg1)), 
                               method = "agresti-coull")[,c("mean", "lower", "upper")]
mpseg2 <- binom::binom.confint(rowSums(seg2), rep(mcmcs, nrow(seg1)),
                               method = "agresti-coull")[,c("mean", "lower", "upper")]

@@

# results
```{r}
colors <- khroma::color("highcontrast")(3)
ci_col <- as.character(colors[3])
colors <- as.character(colors[c(2,1)])

#========plot the emperical CIs around every loci vs the predicted value==============
# plotting the emperical CIs vs the estiamted mean
pseg <- rbind(data.frame(pop = paste0("n = ", n1/2), p = pseg1, snp = 1:nal, af = allele_frequencies),
              data.frame(pop = paste0("n = ", n2/2), p = pseg2, snp = 1:nal, af = allele_frequencies))

@@

f1 <- ggplot(seg[seg$source == "Mathematical",],
             aes(x = af, y = p)) +
  geom_point(aes(color = in_ci)) +
  geom_errorbar(aes(ymax = upper, ymin = lower)) +
  facet_wrap(~pop) +
  theme_bw() +
  theme(strip.background = element_blank()) +
  scale_color_manual(values = colors) +
  guides(color = guide_legend(title = "In CI")) +
  ylab(bquote(italic(P*"("*S[jq]*")"))) +
  xlab("Minor Allele Frequency")

ggsave("Figure1.pdf", f1, width = 14, height = 7)

tapply(seg[seg$source == "Mathematical", "in_ci"], seg$pop[seg$source == "Mathematical"], mean) # percent in interval
#===============plot all of the observed values vs the estimated value for the sums=============
# plot the estimated prediction interval vs the observations and the confint from the simultations
nap_pi <- function(alpha, vseg, n) qnorm(1 - (alpha/2))*sqrt(vseg)*sqrt(1 + (1/n))
nap_ci <- function(alpha, vseg, n) qnorm(1 - (alpha/2))*sqrt(vseg)/sqrt(n)

@@

# CI/PI
msum_ci1 <- nap_ci(.05, var(msum1), length(msum1))
msum_ci2 <- nap_ci(.05, var(msum2), length(msum2))

sum_pi1 <- nap_pi(0.05, sum(vseg1), nal)
sum_pi2 <- nap_pi(0.05, sum(vseg2), nal)

# add to table
msum_tab <- data.frame(nseg = c(mean(msum1), mean(msum2)),
                       lower = c(mean(msum1) - msum_ci1, mean(msum2) - msum_ci2),
                       upper = c(mean(msum1) + msum_ci1, mean(msum2) + msum_ci2),
                       pop = c(paste0("n = ", n1/2), paste0("n = ", n2/2)))

sum_tab <- tapply(seg[seg$source == "Mathematical",]$p, seg[seg$source == "Mathematical",]$pop, sum)
sum_tab <- as.data.frame(sum_tab)
sum_tab$pop <- rownames(sum_tab)
colnames(sum_tab)[1] <- "nseg"
sum_tab <- merge(sum_tab, msum_tab[,c("pop", "lower", "upper")], by = "pop")
sum_tab$in_ci <- sum_tab$nseg <= sum_tab$upper & sum_tab$nseg >= sum_tab$lower

# add pi
all_mcmc_sums <- rbind(cbind.data.frame(nseg = msum1, pop = paste0("n = ", n1/2)),
                   cbind.data.frame(nseg = msum2, pop = paste0("n = ", n2/2)))
pi_tab <- data.frame(pop = c(paste0("n = ", n1/2), paste0("n = ", n2/2)),
                     lower = c(sum_tab$nseg[sum_tab$pop == paste0("n = ", n1/2)] - sum_pi1,
                               sum_tab$nseg[sum_tab$pop == paste0("n = ", n2/2)] - sum_pi2),
                     upper = c(sum_tab$nseg[sum_tab$pop == paste0("n = ", n1/2)] + sum_pi1,
                               sum_tab$nseg[sum_tab$pop == paste0("n = ", n2/2)] + sum_pi2))
all_mcmc_sums <- merge(all_mcmc_sums, pi_tab, by = "pop")
all_mcmc_sums$in_pi <- all_mcmc_sums$nseg <=  all_mcmc_sums$upper & all_mcmc_sums$nseg >=  all_mcmc_sums$lower

Figure2 <- ggplot(sum_tab, aes(x = pop, y = nseg, color = in_ci)) + 
  geom_jitter(data = all_mcmc_sums, aes(x = pop, y = nseg, color = in_pi), height = 0, alpha = .2) +
  geom_point(size = 4, shape = 17) +
  geom_errorbar(data = sum_tab, aes(ymax = upper, ymin = lower, x = pop), 
                inherit.aes = FALSE, width = .25, color = ci_col, linewidth = 1) +
  geom_errorbar(data = pi_tab, aes(ymax = upper, ymin = lower, x = pop), 
                inherit.aes = FALSE, width = .5, linetype = "dashed", color = ci_col, linewidth = 1) +
  theme_bw() +
  theme(strip.background = element_blank()) +
  scale_color_manual(values = colors) +
  guides(color = guide_legend(title = "In CI/PI")) +
  ylab(bquote(italic(E*"("*N[S]*")"))) +
  xlab("Population")

ggsave("Figure2.pdf", Figure2, height = 7, width = 14)

tapply(all_mcmc_sums$in_pi, all_mcmc_sums$pop, mean, na.rm = TRUE)

#==========others==================

@@

# variances, overall
overall_vars <- rbind(cbind.data.frame(estimated = sum(vseg1),
                                       observed = var(msum1),
                                       pop = paste0("n = ", n1/2)),
                      cbind.data.frame(estimated = sum(vseg2),
                                       observed = var(msum2),
                                       pop = paste0("n = ", n2/2)))
```

# Misc
```{r}
# show difference in HWE divergence with different combinations of 5 minor alleles in 100 genotypes
exd <- rbind(c(rep(0, 45), rep(1, 5)),
             c(rep(0, 46), 2, rep(1, 3)),
             c(rep(0, 47), rep(2, 2), 1))
exds <- import.snpR.data(as.data.frame(exd), snp.meta = data.frame(snp = 1:3), mDat = -9)
exds <- calc_hwe(exds)
get.snpR.stats(exds, stats = "hwe")

exc <- counts(exd)
rarefact(exc, 10)
```
